### 2018-05-08 write-up/implementation.tex

@@

Here, I will describe the syntax and semantics.

\section{Elaboration and Inference}

\section{Compiling and Metaprogramming}

>>>>

\prechapter{%
    I implemented LT4LA in OCaml, however I strongly believe the ideas
    described in this chapter can be applied easily to other languages and also
    are modular enough to extend the OCaml implementation to output to different
    back-end languages. I will show how a small core language with a few features
    can be the target of heavy-desugaring of typical linear-algebra programs. This
    core language can then be elaborated and checked for linearity before
    performing some simple and predictable optimisations and emitting (in this
    particular implementation) OCaml code that is not obviously safe and correct
    (with respect to linearity).
}%

\section{Structure of LT4LA}

LT4LA follows the structure of a typical compiler for a (E)DSL\@. From the
start, I made a concerted effort to (1) write pure-functional code (typically
using a monadic-style) which helped immensely with modularity and debugging
when tests showed errors (2) produce readable, useful and precise
error-messages in the hope that someone who did not understand linear types
could still use the LT4LA (3) write tests and set-up
continuous-integration for all non-trivial functions so that I could spot and
correct errors that were not caught by OCaml's type-system whenever I
implemented new features or or refactored my code.

\begin{enumerate}

    \item \textbf{Parsing \& Desugaring}. A generated, LR(1) parser parses a text
        file into a syntax tree, which is then desugared into a smaller, more
        concise abstract syntax tree. The former aims to mimic OCaml syntax
        with a few extensions and keywords so that it is familiar and thus easy
        to pick-up for OCaml users. The latter allows for the type-checker to
        be simpler to implement and easier to specify. In general, this part
        will vary for different languages or can be dealt with differently
        using combinators (the EDSL approach) or a syntax-extension if the host
        language offers such support.

    \item \textbf{Type-checking}. The abstract syntax tree is explicitly-typed,
        with some inference to make it less verbose and more convenient to
        write typical programs.

    \item \textbf{Matrix Expressions}. During type-checking, if a matrix-expression
        is encountered, it is either successfully elaborated into an expression
        in the abstract syntax tree which is then consequently type-checked, or
        fails to find suitable routines to calculate the given expression.

    \item \textbf{Code Generation}. The abstract syntax tree is translated into
        standard OCaml and a few-particular `optimisations' are made to produce
        more readable code. This process is type-preserving: the linear type
        system is embedded into OCaml's type system, and so when the OCaml compiler
        compiles the generated code, it acts as a sanity check on the code produced.

    \item \textbf{Executable Artifacts}. A transpiler and a REPL are the main artifacts
        produced for this thesis.

    \item \textbf{Tests}. As mentioned before, almost all non-trivial functions
        have tests to check their behaviour. The output of the transpiler
        was also tested by having the build system generate code at OCaml code
        which could then be compiled and tested as usual.

\end{enumerate}

@@

A full description of the core language can be found in
Appendix~\ref{chap:ott_spec}. To make a linearly-typed language usable, we need
some way of using values zero or more than once, as we would an intuitionistic
value. For this, we have in the type-expressions the !-constructor and in the
term-expressions the \textbf{many}-constructor. For soundness, I implemented
\emph{value-restriction} for this type system, only \emph{values} (anything not
an array, matrix or an expression that can be reduced) that use \emph{no linear
values} can be marked with !. This means we can freely have integers, booleans
and (appropriate) functions usable as normal. Recursive functions can only use
linear values that are passed in as arguments to the function.

% TODO Once I've implemented the <- syntax, this should just include
% ../test/examples/weighted_avg_infer.lt directly
\begin{figure}[tb]
    \begin{minted}[linenos]{ocaml}
let rec f (!i : !int)
        (!n : !int)
        (!x0 : !elt)
        (write : z arr)
        ('x)
        (weights : 'x arr)
        : z arr * 'x arr =
  if n = i then (write, weights) else
  let !w0 <- weights[0] in
  let !w1 <- weights[1] in
  let !w2 <- weights[2] in
  let !x1 <- write[i] in
  let !x2 <- write[i + 1] in
  let !x : !elt = w0 *. x0 +. w1 *. x1 +. w2 *. x2 in
  f (i + 1) n x1 (write[i] := x) _ weights
    \end{minted}

    \caption{A recursive function to perform a \emph{in-place} (simplified)
        1D-convolution (sliding weighted-average) using a read-only array of
        weights and a writeable array, originally containing the input
        array.}\label{fig:oned_conv}

\end{figure}

So, how does a typical programmer write code that uses a matrix or an array
more than once? Figure~\ref{fig:oned_conv} shows an example program that
executes a (simplified) 1D-convolution (sliding weighted-average) using a
read-only array of three weights. On line 1, \ltfla{let rec} introduces a
recursive function `f', followed by explicitly annotated formal parameters.
Intuitively, \ltfla{!id} allows \ltfla{id} to be used intuitionistically. The
types \ltfla{!int} and \ltfla{!elt} correspond to integers and element types
(for example, 64-bit floating-point).

Array and matrix types are parameterised by \emph{fractional-capabilities}.  A
fraction of 1 ($2^0$) represents complete ownership of a value; in
particular, this allows a programmer to write or free it. Creating an array
gives you ownership of it; the function \ltfla{array : !int --o z arr}
(where \ltfla{z} represents `0'). Once you have ownership of an
array/matrix, you can free it: \ltfla{free : z arr --o unit}.
Importantly, because a linear-value may only be used once, the array just freed
is \emph{out of scope} for following expressions, preventing use-after-free.
Ownership also enables you to write to the array: \ltfla{set : z arr
--o !int --o !elt --o z arr} (the syntax \ltfla{w[i] := j} is just
sugar for \ltfla{set w i j}). Here, linearity prevents accessing values
which represented the array \emph{before} the mutation.

Any fraction less than 1 (for simplicity, limited to $2^{-k}$ in this system,
for a positive integer $k$) represents read-only access. So, the \ltfla{'x}
represents a natural number (either a zero \ltfla{z}, variable \ltfla{'x} or a
successor (+1) of a natural number). Hence, you can read from (index) any array
\ltfla{get : 'x .  'x arr --o !int --o !elt * 'x arr} (the syntax \ltfla{let !v
<- w[i]} is just sugar for \ltfla{let (!v, w) = get _ w i}). Transparently
rebinding \ltfla{w} with the returned value means a program can appear to use
\ltfla{w} multiple times; the underscore is how a programmer tells the compiler
to automatically \emph{infer} the correct fractional-capability based on the
other arguments passed to the function. Fractions also permit aliasing:
\ltfla{share : 'x . 'x arr --o 'x s arr * 'x s arr}. However, given the types
of the primitives provided, these aliases can only be read from and not written
to. If you want to write to this array, you must \ltfla{unshare: 'x . 'x s arr
--o 'x s arr --o 'x arr} it until you are left with a value of type \ltfla{z
arr}, guaranteeing no other aliases exists.

Given this set-up, we now \emph{statically} have \emph{perfect} information
about aliasing and ownership of values in the program. We can only write to an
array when we own it; ownership guarantees no other aliases exist in scope at
the point of usage. In the next section, I show how this perfect information
can be used to write more natural-looking code using value-semantic expressions
which behave in precisely the way we intend it to. Now the programmer need not
resort to manually tracking and inserting \texttt{noalias} annotations; instead
they can let the loyal and tireless compiler do the heavy lifting.

\section{Matrix Expressions}

We have now arrived at the titular \emph{applications} of linear types. I will
show how, in addition to preventing the errors explained in the previous
section, we can take linear types one step further and apply it to the domain
of efficient compilation of matrix-expressions.

\begin{figure}[tb]
    \inputminted[linenos]{fortran}{kalman.f90}
    \caption{Kalman filter in Fortran 90 (Credit: Matthew
        Rocklin~\cite{rocklin_gist}).}\label{fig:fortran_kalman}
\end{figure}

In Figure~\ref{fig:fortran_kalman}, we see the difficulty of efficiently
implementing a \emph{Kalman filter}, a powerful and set of equations applicable
to a wide variety of problems. From the comments, we see that every variable is
annotated with the step/matrix expression that it will hold at some point
during the computation (an equivalent alternative, say in C++, could be to have
a meaningful name for each step/matrix expression and manually annotate/keep
track of which names alias the same location).

In contrast, Figure~\ref{fig:ltfla_kalman}, offers the advantages of
\begin{itemize}
    \item aliasing: labelling each step with a different, more meaningful variable name,
    \item easily spotting which resources are being passed in and which are
        allocated for the function (new/copy),
    \item unambiguously seeing \emph{when} and what values are freed;
\end{itemize}
and have the compiler automatically ensure the safety of each of the above by respectively
\begin{itemize}
    \item making it impossible to refer to steps/values which are no longer usable,
    \item ensuring all values are declared and \emph{initialised} correctly before they are used,
    \item checking no values are used after they are freed \emph{and} ensuring no values are leaked.
\end{itemize}

Indeed, an inexperienced programmer could take the  na\"ive approach of just
copying sub-expressions by default and then letting the compiler tell it which
copies are never used and removing them systematically until it type checks.
While it is not quite a black-box, push-button compilation of an expression, I
would argue that, it is just as easy (if not easier) to become familiar with as
Rust and its borrow-checker.

\begin{figure}[tb]

    \begin{minted}[linenos]{ocaml}
let !kalman
        ('s) (sigma : 's mat) (* n, n *)
        ('h) (h : 'h mat)     (* k, n *)
        ('m) (mu : 'm arr)    (* n    *)
        (r_1 : z mat)         (* k, k *)
        (data_1 : z arr)      (* k    *)
        : ('s mat * ('h mat * 'm arr)) * (z arr * z mat) =
    let ((!k, !n), h) = dimM _ h in
    let zeros = zeroM n n in
    let sigma_h = new [| sym (up, sigma) * h |] in
    let r_2 = [| h * sigma_h + r_1 |] in
    let mu_1 <- copy mu in
    let sigma_1 <- copyM sigma in
    let data_2 = [| h * mu + -1. * data_1 |] in
    let h_1 <- copyM h in
    let h_sol = [| sym (up, r) ^ -1 * h_1 ] in
    let data_sol = [| sym (up, r) ^ -1 * data_2 |] in
    let h_h1 = new [| h * h_sol |] in
    let h_data = new [| h * data |] in
    let h_h1_sigma = new [| sym (up, h_h1) * sigma |] in
    let mu_2 = [| sym (up, sigma) * h_data * mu_1 |] in
    (* non-linear use of sigma_c..? *)
    let sigma_c = [|  -1. * sym (up, sigma_c) * h_h1_sigma + sigma_c |] in
    (* frees... *)
    (mu_c , sigma_c)
    \end{minted}

    \caption{Kalman filter in LT4LA.}\label{fig:ltfla_kalman}

\end{figure}

\subsection{TO DO: Elaboration}

So-called `matrix \emph{expressions}' are still `side-effecting'/consuming
linear values, and do not produce \emph{values} but a \emph{sequence of
(re-)bindings} which in turn dictate what values are still available/in-scope
after the computation.  As such, compilation cannot be done purely
compositionally via structural-induction on the expression language; it is most
concisely expressed via CPS, where the result of elaborating a matrix
expression is a function that takes as its argument the rest of the computation
expecting to use the result it has just made available under that requested
name.

\subsection{TO DO: Inference}

\section{TO DO: Code Generation}

High-level overview: primitives, translations, optimisations, difficulty, build system.

\subsection{TO DO: Compiling Constructs to OCaml}

\subsection{TO DO: Metaprogramming}

\section{TO DO: Theory}

Fractional permissions, not proven soundness etc\. but compilation preserves
types as checked by OCaml.

### 2018-05-08 write-up/introduction.tex
Start background and add matrices to semantics

\chapter{Introduction}

@@

>>>>

\chapter{TO DO: Introduction}

@@

\prechapter{%
    In this thesis, I will argue that linear types are an appropriate,
    \emph{type-based formalism} for the problem of \emph{efficient}
    matrix-expression compilation. I will show that framing the problem using
    linear types can help \emph{reduce bugs} by making precise and explicit the
    informal, ad-hoc practices typically employed by human experts and linear
    algebra \emph{compilers} and automate checking them. As evidence for this
    argument, I will show programs written with this safety, precision and
    explicitness (1) can be just as pleasant and convenient for a programmer as
    less efficient, but higher-level linear algebra libraries and (2) perform
    just as \emph{efficiently and predictably} as lower-level, less readable
    and more error-prone linear algebra libraries.
}%

### 2018-05-08 write-up/related_work.tex
Start background and add matrices to semantics

\chapter{Related Work}

\begin{guidance}

    This chapter covers relevant (and typically, recent) research
    which you build upon (or improve upon). There are two complementary
    goals for this chapter:
    \begin{enumerate}
      \item to show that you know and understand the state of the art; and
      \item to put your work in context
    \end{enumerate}

    Ideally you can tackle both together by providing a critique of
    related work, and describing what is insufficient (and how you do
    better!)

    The related work chapter should usually come either near the front or
    near the back of the dissertation. The advantage of the former is that
    you get to build the argument for why your work is important before
    presenting your solution(s) in later chapters; the advantage of the
    latter is that don't have to forward reference to your solution too
    much. The correct choice will depend on what you're writing up, and
    your own personal preference.

\end{guidance}

>>>>

\chapter{Related Work}\label{chap:related}

% \begin{guidance}
% 
%     This chapter covers relevant (and typically, recent) research
%     which you build upon (or improve upon). There are two complementary
%     goals for this chapter:
%     \begin{enumerate}
%       \item to show that you know and understand the state of the art; and
%       \item to put your work in context
%     \end{enumerate}
% 
%     Ideally you can tackle both together by providing a critique of
%     related work, and describing what is insufficient (and how you do
%     better!)
% 
%     The related work chapter should usually come either near the front or
%     near the back of the dissertation. The advantage of the former is that
%     you get to build the argument for why your work is important before
%     presenting your solution(s) in later chapters; the advantage of the
%     latter is that don't have to forward reference to your solution too
%     much. The correct choice will depend on what you're writing up, and
%     your own personal preference.
% 
% \end{guidance}

\prechapter{%
    Now that I have described my contributions, I will explain how it relates
    to existing work, leaving brief discussions on future work to the next
    chapter. I strongly believe matrix expression compilation research would
    benefit greatly from a comprehensive literature review but unfortunately
    that is beyond the scope of this chapter.
}%

\section{Matrix Expression Compilation}

% For decades, small gains in efficiency of \emph{executing} numerically
% intensive code have and continue to translate into massive savings in costs and
% time. This has spurned mountains of research into an almost bewildering variety
% of hardware, software and compilation techniques, typically entrusted with a
% (team of) domain-specific expert programmer(s) to apply in real world
% situations.
% 
% Given the rise of big data and machine learning, there is now a movement to
% improve \emph{programmer} efficiency in this domain after focusing for so long
% on \emph{computational} efficiency. Matrix expression compilers promise to
% generate code that is reasonably close to `optimal'/what a domain-expert may
% write but without the need to acquire years of experience and depth of
% knowledge (a similar trade-off to garbage-collection for memory management or
% software-transactional memory for concurrency).

Most of the projects below try to be fully-automated black-boxes which model
computing a matrix expression as some sort of graph with informal, ad-hoc rules
about what can and should be copied or modified in-place. Allocations,
temporaries and common sub-expressions are details invisible to the programmer,
left to the compiler.

The matrix expression `compiler' as implemented in LT4LA is intended to be a
mere proof-of-concept of how linear types can and arguably should be used
\emph{ergonomically}. I have taken the approach of attempting to help
programmers precisely and explicitly capture, using types, the practices
prevalent in code they already write.

I believe the advantages of my approach are two-fold (1) \emph{predictable}
performance and (2) more accurate modelling of how low-level kernels handle
their resources. My confidence in the latter claim comes from finding at least
two errors in the Fortran code output by SymPy to compute a Kalman filter, one
to do with aliasing (resulting in code that was not legal Fortran) that would
not have type-checked had it been translated via LT4LA as an intermediate
representation (the other error to be explained later
in~\ref{subsec:lin_dep_types}).

\subsection{SymPy}

SymPy is a symbolic computer algebra system for Python; its matrix expression
compiler~\cite{rocklin_thesis} uses a term-rewrite system, with rules supplied
by a BLAS expert (which must be strongly-normalising, that is, never cause a
loop) but need not be confluent (there can be more than one solution per
expression). Rules include expressions to match on, the expression it can
produce, information about the expressions (such as whether the matrix is
symmetric or full-rank) and information about which variable is updated
in-place.

\subsection{Clak and Cl1ck}

Clak and Cl1ck~\cite{fabregat_thesis} were developed independently around the
same time as SymPy's matrix expression compilation. Clak attempts to produce
\emph{multiple} algorithms for a single matrix expression, by considering a
wider matrix expression grammar and more matrix properties and inference rules.
These algorithms assume basic building blocks such as products and
factorisations. Cl1ck attempts to take on the challenge of writing BLAS/LAPACK
like libraries too, by generating lower-level loop-based blocked routines for
the aforementioned basic building blocks, in the spirit of the
FLAME~\cite{gunnels_flame} project.

\subsection{Linnea and Taco}

Linnea~\cite{linnea} and Taco~\cite{taco} are two newer contenders to Clak and
Cl1ck respectively. Linnea continues the work of Clak to producing real
executable code for \emph{existing} libraries and kernels, as well as
incorporating work on a \emph{generalised} matrix chain algorithm~\cite{gmc}.
Taco (\emph{Tensor Algebra COmpiler}) focuses on emitting efficient routines
for expressions in tensor index notation, with many optimisations for
\emph{sparse} tensors.

\section{Metaprogramming}

Most of the compilers in the aforementioned projects usually built, analysed,
compiled, ran (and in some cases, dynamically loaded) expressions (including
functions) at runtime, similar to how regular expressions are handled in most
languages -- in particular, even when the regular expression is known at
compile time.

In LT4LA, I took the approach of having a concrete syntax and expression
language which was then translated and made available as a \emph{typed
expression} to other modules \emph{at compile time} via the build system. Apart
from convenience in programming and testing, there was nothing inherent in the
approach that prevented me from using OCaml's PPX syntax-extensions so that I
could write normal OCaml expressions from within OCaml and have them checked
for linearity before compilation.

Having a statically compiled language and a build system as so affords the
advantage of eliminating the runtime overheads mentioned at the start of this
section. However, there is some useful information (such as matrix dimensions
for the matrix chain algorithm) which is sometimes known only at runtime (but
once known, usually fixed). In these cases, using \emph{multi-stage
programming} would be a better approach to implementing a matrix expression
compiler.

\subsection{MetaOCaml and Scala LMS}

MetaOCaml~\cite{metaocaml} and Scala with Lightweight Modular
Staging~\cite{scala_lms} are systems which support multi-stage programming. A
typical example of this is the generation of Fast Fourier-Transform kernels,
specialised to a desired array length.

\subsection{Expression Templates in C++ Libraries}

Expression templates are a commonly used compile-time metaprogramming
technique, used by Eigen, uBLAS and Armadillo to name a few. If known at
compile-time, matrix dimensions can also be passed in as template arguments to
ensure operations match (otherwise checking at runtime). In Eigen, such
features are combined with heuristics to enable \emph{lazy evaluation} and
automatically determine whether a sub-expression is evaluated into a temporary
variable or not.

They perform rudimentary pattern-matching and in some cases, loop-fusion, to
avoid evaluating expressions in a purely binary manner (invoking the bane of a
C++ programmer: temporaries and unnecessary copies) when possible (either by
translating to a library kernel call or, as an example, inlining a $v := a + b
+ c$ vector expression into one loop).

As is the case with LT4LA, this approach shares \emph{some} elimination of
runtime overheads, but not all, thanks to the heuristics surrounding lazy
evaluation and evaluating sub-expressions into temporaries. This comes at the
cost of a user being able to easily inspect the generated C++ code, losing
explicitness.

\section{Types}

Apart from lazy evaluation, the following projects show how instead of a (E)DSL
library approach, we could have type-level resource management provided far
more conveniently and naturally at the \emph{language} level. The difference is
that a library can be designed, shipped and used now whereas language features
take time and can have unintended interactions with other language features. My
hope is that once people are convinced of the utility type-level resource
management by using a library, the impetus for integrating such features into
the language follows.

\subsection{Lazy Evaluation}

A particularly strong advantage LT4LA has over other libraries that use lazy
evaluation is, funnily enough, linear types, more precisely and the static and
perfect information they guarantee about aliasing: with Owl, every graph-node
has only one node in or out; with Eigen, every ``assignment'' has a
\texttt{noalias} annotation.

This simplifies the rules and exceptions a programmer reasoning about memory
usage needs to remember. Of course, now the programmer has to figure out how
they are using their temporaries, but because matrices are linearly-typed,
redundant copies/missed frees can be pointed out by the compiler, guiding them
towards a satisfying solution.

\subsection{Futhark}

Futhark~\cite{futhark} is a second-order (meaning it supports functions such as
map and fold/reduce) array combinator (meaning array operations can be fused
into streams to reduce temporaries) language designed for efficient parallel
compilation. It supports ML-style modules, loops, limited parametric
polymorphism, size types and uniqueness types. Intuitively, where linearity
provides a local guarantee of ``this value is not aliased in this scope'',
uniqueness types provide a global guarantee of ``this value is not aliased
anywhere''. Its combinators are more expressive than typical linear algebra
library kernels so encourages shorter, more declarative linear algebra code.

\subsection{Substructural Features in Rust}

Rust~\cite{rust} is a (relatively) new systems programming language aiming to
bring the last two decades of programming language research to the masses in a
usable and friendly manner. Its \emph{borrow-checker} is the feature most
relevant to this thesis because it statically attempts to prevent many
resource-related bugs. Although there are a few linear algebra libraries for
Rust under development, careful use of its macro system and borrow-checker
could make it the safest and easiest to use language for linear algebra
projects to come. Its struggle is more likely to be against the inertia of the
large amounts of C++/Fortran code already out there rather than its usability
or benefits.

\subsection{Linear Types in Haskell}

Linear types have been incorporated into a branch of the Glasgow Haskell
Compiler~\cite{retrofitting} in an attempt to provide safe, functional
streaming and IO (after people saw the potential from libraries providing
linearity features). Practical benefits include zero-copy buffers and
eliminating garbage collection in certain situations by allowing the user to
safely manage memory. The fact that it can and has been done gives me hope that
other languages will also see the value and adopt some form of
resource-management in their type systems.

\subsection{Linear and Dependent Types in Idris}\label{subsec:lin_dep_types}

Dimension mismatches are seen as an irritating but small inconvenience when
writing linear algebra code. However, the second error I found in the Fortran
code output by SymPy to compute a Kalman filter was a dimension/transposition
error. Although we would not need full dependent types to solve dimension
mismatches (symbolic size types would be sufficient), managing properties about
matrices could be done at the type-level in a dependently typed setting.

We could then express the usual properties and results of operations at the
type-level, ensuring, for example, that certain functions are called only when
the matrix is symmetric and can be written to. Idris (a Haskell inspired
language with dependent types) has had experimental support for uniqueness
types since its early days and now a linear types extension~\cite{idris_linear}
is also being worked on based on new research around integrating linear and
dependent types~\cite{atkey}.
